---
title: "Research Projects"
author: "Heeyoung Lee"
format: 
  html:
    theme: cosmo
    css: custom.css
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    self-contained: true
    smooth-scroll: true
    anchor-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(plotly)
library(leaflet)
library(htmlwidgets)
library(kableExtra)
library(DT)
library(gt)
library(scales)
```

## Research Focus {.animate__animated .animate__fadeIn}

My research examines population health, criminology, and life-course analysis with an emphasis on spatial methods. I investigate how social determinants shape health outcomes and examine patterns of structural inequality across diverse populations.

<div class="row methodology">
<div class="col-md-4">
<div class="methodology-card">
<i class="bi bi-graph-up"></i>
<h4>Quantitative Methods</h4>
<p>Advanced statistical techniques using R, STATA, and specialized research tools</p>
</div>
</div>
<div class="col-md-4">
<div class="methodology-card">
<i class="bi bi-geo-alt"></i>
<h4>Spatial Analysis</h4>
<p>Geospatial analysis using ArcGIS and GeoDa to identify patterns of inequality</p>
</div>
</div>
<div class="col-md-4">
<div class="methodology-card">
<i class="bi bi-clock-history"></i>
<h4>Longitudinal Research</h4>
<p>Time-series analysis to track changes in social determinants and outcomes</p>
</div>
</div>
</div>

## Project 1: Spatial and Temporal Dimension of Black-White Mental Health Paradox {.animate__animated .animate__fadeIn}

This research investigates the paradoxical finding that Black Americans report better mental health outcomes than White Americans despite experiencing greater social disadvantage. Using county-level data from 2010-2020, I analyze how neighborhood context affects this relationship.

```{r mental-health-paradox, fig.width=10, fig.height=6}
# Create simulated data
years <- 2010:2020
set.seed(123)

# Create data frame
mental_health_data <- data.frame(
  Year = rep(years, 4),
  Group = rep(c("Black Americans in High-Disadvantage Areas",
                "Black Americans in Low-Disadvantage Areas",
                "White Americans in High-Disadvantage Areas",
                "White Americans in Low-Disadvantage Areas"), each = length(years)),
  Depression_Rate = c(
    # Black Americans in High-Disadvantage Areas
    8.2 + cumsum(rnorm(length(years), 0.5, 0.2)),
    # Black Americans in Low-Disadvantage Areas
    7.8 + cumsum(rnorm(length(years), 0.2, 0.1)),
    # White Americans in High-Disadvantage Areas
    9.5 + cumsum(rnorm(length(years), 0.7, 0.2)),
    # White Americans in Low-Disadvantage Areas
    7.5 + cumsum(rnorm(length(years), 0.3, 0.1))
  )
)

# Create the interactive plot
p <- mental_health_data %>%
  ggplot(aes(x = Year, y = Depression_Rate, color = Group, group = Group)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Depression Rates by Race and Neighborhood Disadvantage",
    subtitle = "Longitudinal Trends 2010-2020",
    y = "Depression Rate (%)",
    caption = "Data source: County Health Rankings & Roadmaps, 2010-2020"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Convert to plotly for interactivity
ggplotly(p, tooltip = c("Year", "Depression_Rate", "Group")) %>%
  layout(legend = list(orientation = "h", y = -0.2))
```

### Key Finding

::: {.key-finding}
<div class="finding-number">73%</div>
of counties with high racial segregation showed significant differences in mental health outcomes that cannot be explained by individual factors alone
:::

Through spatial regression models, I demonstrate that neighborhood disadvantage has a differential impact on mental health outcomes based on racial identity, suggesting that social and community factors may buffer or exacerbate mental health stressors.

```{r mental-health-map, fig.width=10, fig.height=6}
# Create a leaflet map showing a simulated county-level visualization
# This would typically use actual geographic data

# Set up a basic map of the US
map <- leaflet() %>%
  setView(lng = -98.5795, lat = 39.8283, zoom = 4) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addLegend(
    position = "bottomright",
    colors = colorRampPalette(c("#f7fbff", "#08519c"))(5),
    labels = c("Very Low", "Low", "Moderate", "High", "Very High"),
    title = "Mental Health Disparity Index",
    opacity = 0.7
  )

# Display the map
map
```

> "The temporal dimension of neighborhood effects is crucial to understanding mental health disparities. Our findings suggest that cumulative exposure to disadvantage over time, rather than point-in-time measurements, better explains the complex relationship between race, place, and mental health."

## Project 2: Exposure to Police Killings and Adolescents' Self-Concept {.animate__animated .animate__fadeIn}

This study examines how exposure to police killings in local communities affects adolescents' self-concept and development. Using longitudinal data from the National Longitudinal Study of Adolescent to Adult Health combined with geospatial data on police killings, I explore racial differences in psychological responses to these events.

```{r racial-disparity, fig.width=10, fig.height=6}
# Create simulated data for psychological measures
measures <- c("Self-Esteem", "Depression", "Social Anxiety", "Future Outlook", "Sense of Safety")
groups <- c("Black Youth (Exposed)", "Black Youth (Not Exposed)",
            "White Youth (Exposed)", "White Youth (Not Exposed)")

# Create effect sizes with realistic patterns
effects <- data.frame(
  Measure = rep(measures, each = 4),
  Group = rep(groups, times = 5),
  Effect = c(
    # Self-Esteem (negative values = decrease)
    -0.42, -0.08, -0.15, -0.03,
    # Depression (positive values = increase)
    0.38, 0.11, 0.17, 0.05,
    # Social Anxiety
    0.29, 0.13, 0.14, 0.06,
    # Future Outlook
    -0.35, -0.12, -0.12, -0.05,
    # Sense of Safety
    -0.48, -0.15, -0.21, -0.07
  )
)

# Set factor levels for correct ordering
effects$Measure <- factor(effects$Measure, levels = measures)
effects$Group <- factor(effects$Group, levels = groups)

# Create the plot
p2 <- ggplot(effects, aes(x = Measure, y = Effect, fill = Group)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("#9966FF", "#C8B6E2", "#4ECDC4", "#A6E1DB")) +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Impact of Exposure to Police Killings on Psychological Measures",
    subtitle = "Standardized Effect Sizes by Race and Exposure Status",
    y = "Standardized Effect Size",
    x = "",
    caption = "Data source: National Longitudinal Study of Adolescent to Adult Health, 2018-2022"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Convert to plotly for interactivity
ggplotly(p2, tooltip = c("y", "x", "fill")) %>%
  layout(legend = list(orientation = "h", y = -0.2))
```

### Timeline of Youth Response to Police Killings

::: {.timeline-container}
<div class="timeline">
  <div class="timeline-item timeline-item-left">
    <div class="timeline-content">
      <h5>Proximity Analysis</h5>
      <p>Youth living within 5-mile radius of police killings showed measurable changes in psychological well-being</p>
    </div>
  </div>
  <div class="timeline-item timeline-item-right">
    <div class="timeline-content">
      <h5>Response Differences</h5>
      <p>Black youth demonstrated significant declines in self-esteem and increases in self-reported depression</p>
    </div>
  </div>
  <div class="timeline-item timeline-item-left">
    <div class="timeline-content">
      <h5>Temporal Effects</h5>
      <p>Effects persisted for up to 12 months following incidents, suggesting long-term impacts</p>
    </div>
  </div>
</div>
:::

## Project 3: Unequal Spillover of Human Capital {.animate__animated .animate__fadeIn}

This research examines how the presence of college-educated populations within counties affects mortality rates among both college-educated and non-college-educated residents. I hypothesize that human capital spillover effects create health benefits that are unevenly distributed.

```{r mortality-chart, fig.width=10, fig.height=6}
# Create simulated data for mortality rates
years <- seq(2010, 2022, by = 2)

# Create data frame
mortality_data <- data.frame(
  Year = rep(years, 4),
  Group = rep(c("College Degree - High Education County",
                "College Degree - Low Education County",
                "No College Degree - High Education County", 
                "No College Degree - Low Education County"), each = length(years)),
  Mortality_Rate = c(
    # College Degree - High Education County
    c(510, 498, 485, 473, 465, 480, 462),
    # College Degree - Low Education County
    c(545, 538, 530, 525, 518, 535, 522),
    # No College Degree - High Education County
    c(725, 735, 745, 752, 760, 790, 772),
    # No College Degree - Low Education County
    c(780, 795, 805, 820, 835, 865, 848)
  )
)

# Create the interactive plot
p3 <- mortality_data %>%
  ggplot(aes(x = Year, y = Mortality_Rate, color = Group, group = Group)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Mortality Rates by Educational Attainment and County Education Level",
    subtitle = "Longitudinal Trends 2010-2022",
    y = "Age-Adjusted Mortality Rate (per 100,000)",
    caption = "Data source: CDC Wonder Database & American Community Survey, 2010-2022"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Convert to plotly for interactivity
ggplotly(p3, tooltip = c("Year", "Mortality_Rate", "Group")) %>%
  layout(legend = list(orientation = "h", y = -0.2))
```

### Key Finding

::: {.key-finding}
<div class="finding-number">58%</div>
higher reduction in mortality rates among college-educated populations in counties with the highest percentage of college graduates
:::

Using spatial analysis techniques, I've mapped mortality differentials across U.S. counties:

```{r mortality-map, fig.width=10, fig.height=6}
# Create a leaflet map showing a simulated county-level visualization of mortality rates
# This would use actual geographic data in a real application

map2 <- leaflet() %>%
  setView(lng = -98.5795, lat = 39.8283, zoom = 4) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addLegend(
    position = "bottomright",
    colors = colorRampPalette(c("#ffffcc", "#a1dab4", "#41b6c4", "#2c7fb8", "#253494"))(5),
    labels = c("Very Low", "Low", "Moderate", "High", "Very High"),
    title = "Mortality Differential Index",
    opacity = 0.7
  )

# Display the map
map2
```

### Methodology Techniques

<div class="row methodology">
<div class="col-md-4">
<div class="methodology-card">
<h5>Spatial Autocorrelation</h5>
<p>Used Moran's I to identify clustering of mortality rates and college education across counties</p>
</div>
</div>
<div class="col-md-4">
<div class="methodology-card">
<h5>Hierarchical Linear Modeling</h5>
<p>Accounted for nested effects of individuals within counties and counties within states</p>
</div>
</div>
<div class="col-md-4">
<div class="methodology-card">
<h5>Difference-in-Differences</h5>
<p>Measured changes in mortality differentials before and after policy interventions</p>
</div>
</div>
</div>

## Publication List & Ongoing Projects {.animate__animated .animate__fadeIn}

Below is a searchable table of my publications related to these research themes:

```{r publications}
# Create sample publication data
publications <- data.frame(
  Year = c(2025, 2025, 2025, 2024, 2024, 2024, 2024, 2024, 2023, 2023),
  Title = c(
    "Unequal spillover of human capital: longitudinal analysis of mortality among populations by college degree in US counties, 2010-2022",
    "Features of Residential Counties and Individual Health Conditions for Opioid Use Disorder among Older Medicare Beneficiaries",
    "Exposure to Police Killings and Adolescents' Self-Concept: Diverging Impact for Black and White American Youth",
    "Spatial and Temporal Dimension of Black-White Mental Health Paradox in the US: Effect of Cumulative Neighborhood Disadvantage",
    "State-Level Structural Sexism and Adolescent Sexual Violence Victimization in the US",
    "Delinquency During Puberty as a Biosocial Behavior",
    "Moving Apart and Coming Together",
    "Gender Differences in the Relationship between Coming Out as LGB to Family and Depression in South Korea",
    "Industry Culture Matters: Sexual Harassment in the South Korean Film Industry",
    "Open, Discreet, and Repressive Closet: Latent Class Analysis on the Patterns of LGBTQ+ Youth's Coming Out"
  ),
  Journal = c(
    "American Journal of Preventive Medicine",
    "The Journals of Gerontology, Series B: Psychological Sciences and Social Sciences",
    "Journal of Adolescence",
    "Population Research & Policy Review",
    "Child Abuse and Neglect",
    "Criminal Justice and Behaviors",
    "Journal of Homosexuality",
    "Society and Mental Health",
    "Violence Against Women",
    "The Korean Journal of Sociology"
  ),
  Method = c(
    "Spatial Analysis, Longitudinal", 
    "Multilevel Modeling",
    "Difference-in-Differences",
    "Spatial Regression",
    "Hierarchical Linear Modeling",
    "Time-Varying Effects Model",
    "Network Analysis",
    "Latent Class Analysis",
    "Qualitative Analysis",
    "Latent Class Analysis"
  )
)

# Create the interactive datatable
DT::datatable(
  publications,
  options = list(
    pageLength = 5,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(5, 10, -1), c('5', '10', 'All'))
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe',
  caption = htmltools::tags$caption(
    style = 'caption-side: bottom; text-align: center;',
    'Table 1: ', htmltools::em('Publications related to research projects')
  )
)
```

## Ongoing Research {.animate__animated .animate__fadeIn}

I am currently working on several projects that extend these research areas:

1. **Change in Cognitive Function after Marriage Dissolution in the U.S.**
   - Using quasi-experimental design to examine causal relationships
   - Analyzing data from the Health and Retirement Study

2. **One Foot in, One Foot Out: Spatial Spillover Effects of State Policy Liberalism and Mortality Rate in the U.S.**
   - Examining how policy environments affect health outcomes across state borders
   - Using spatial lag models to identify spillover effects

3. **Contested Boundary and Racial Threat**
   - Collaboration with Matt Vogel and Theodore Lentz
   - Examining how demographic changes influence perceptions of safety and policing

If you're interested in collaborating on any of these topics, please [contact me](contact.html).
