---
title: "Understanding Difference-in-Differences"
author: "Heeyoung Lee"
date: "Nov 29, 2025"
format: 
  html:
    theme: cosmo
    css: custom.css
    page-layout: full
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 5
    toc-location: left
    self-contained: true
    fig-width: 8
    fig-height: 6
    fig-dpi: 300
    smooth-scroll: true
    highlight-style: github
execute:
  warning: false
  message: false
---

## Understanding Difference-in-Differences

## 1. Introduction - The Cholera Mystery: A Historical Prelude to Causal Inference

In 1854, London faced a devastating cholera outbreak in Soho. The prevailing theory blamed "miasma"—bad air. But physician John Snow suspected contaminated water. He mapped cholera deaths and noticed clustering around the Broad Street water pump. Yet correlation wasn't causation: did the pump cause cholera, or did people near the pump share other risk factors?

Snow found his natural experiment: two water companies served the same neighborhoods. The Lambeth Company had recently moved its water intake upstream (away from sewage), while Southwark & Vauxhall continued drawing contaminated water. Crucially, households on the same street received water from different companies based on historical contracts—not health or wealth.

![**Figure 1.** Water Supply of the Southwark and Vauxhall and Lambeth Companies, London, 1855. This map shows neighborhoods supplied by two different water companies. People consuming water from the Southwark and Vauxhall Company (blue) became ill with cholera, while those consuming water from the Lambeth Company (pink) did not. Thus, the water supply, not social class or behavior, caused the disease. (source: https://www.cooperhewitt.org/2021/12/10/john-snows-cholera-maps/)](C:/Users/User/OneDrive/Github Desktop/idlhy0218.github.io/images/cholera map.jpg)

```{r library-imports, message=FALSE, warning=FALSE}
library(HistData)
library(fixest)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(tidyr)
```

```{r cholera-comparison, fig.cap="**Figure 2.** Snow's natural experiment comparing cholera deaths by water source. Households served by different companies in the same neighborhoods experienced dramatically different mortality rates. Data based on Snow (1855)."}
# Simulate Snow's comparison (simplified)
set.seed(123)

cholera_data <- data.frame(
  company = rep(c("Lambeth (Clean)", "Southwark and Vauxhall (Contaminated)"), each = 100),
  deaths_per_10k = c(rnorm(100, mean = 37, sd = 8),
                     rnorm(100, mean = 315, sd = 40))
)

ggplot(cholera_data, aes(x = company, y = deaths_per_10k, fill = company)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = c(37, 315), linetype = "dashed", alpha = 0.5) +
  scale_fill_manual(values = c("Lambeth (Clean)" = "steelblue", 
                                "Southwark and Vauxhall (Contaminated)" = "coral")) +
  labs(
    title = "Snow's Natural Experiment: Deaths by Water Source",
    subtitle = "Same neighborhoods, different water companies",
    x = "Water Company",
    y = "Deaths per 10,000 Households",
    caption = "Based on Snow (1855)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Snow's insight:** By comparing similar households with different water sources, he isolated the causal effect. Deaths: Lambeth = 37 per 10,000; S&V = 315 per 10,000. The pump was removed, the outbreak ended.

**The challenge:** Snow had a simple before-after comparison at one point in time. But what if we want to study interventions that roll out gradually? What if we can't randomly assign treatment?

### Enter Difference-in-Differences

DiD extends Snow's logic to **over-time comparisons**. Instead of comparing groups at one time, we compare how groups change differently after an intervention. This addresses a critical weakness: even if groups differ at baseline, we can still estimate causal effects if they would have trended similarly without treatment.

**The core insight:** 

- Snow: Compare across groups (spatial variation)
- DiD: Compare across groups AND time (spatiotemporal variation)

By "differencing out" both baseline differences and common trends, DiD isolates the treatment effect—just as Snow isolated water quality by holding neighborhoods constant.

## 2. What Is Difference-in-Differences?

### The Core Idea

DiD estimates treatment effects by comparing how outcomes change over time in two groups: one that receives an intervention (treatment group) and one that doesn't (control group). The method removes two major sources of bias:

1. **Baseline differences** between areas that stay constant over time (first difference)
2. **Common epidemic trends** that affect both areas equally (second difference)

By subtracting out these two sources of bias, we isolate the true causal effect of the pump removal.

### Mathematical Model

The basic 2×2 DiD setup (two groups, two time periods) is:

$$Y_{it} = \beta_0 + \beta_1 \text{Treat}_i + \beta_2 \text{Post}_t + \beta_3 (\text{Treat}_i \times \text{Post}_t) + \epsilon_{it}$$

Where:

- $Y_{it}$ = cholera deaths in district $i$ at week $t$
- $\text{Treat}_i$ = 1 if district near Broad Street pump, 0 if not
- $\text{Post}_t$ = 1 if after pump removal, 0 if before
- $\text{Treat}_i \times \text{Post}_t$ = interaction term
- $\beta_3$ = **DiD estimator** (lives saved by pump removal)

> **Intuition:** Think of this as a regression with three building blocks. $\beta_1$ captures the fact that Broad Street areas were always riskier. $\beta_2$ captures the natural epidemic decline that happened everywhere. The interaction term $\beta_3$ asks: *above and beyond* those two things, did treated areas change more? That extra change is the pump removal effect.

### The DiD Estimator

The DiD estimator calculates:

$$\hat{\beta}_{DiD} = (\bar{Y}_{\text{Broad St,post}} - \bar{Y}_{\text{Broad St,pre}}) - (\bar{Y}_{\text{other,post}} - \bar{Y}_{\text{other,pre}})$$

In plain language:

- **First difference**: How deaths changed near Broad Street pump
- **Second difference**: How deaths changed in other areas
- **DiD effect**: Extra mortality reduction near Broad Street beyond the natural epidemic decline

> **Intuition:** The first difference alone is misleading—deaths near the pump fell partly because of the pump removal *and* partly because cholera was naturally subsiding. The second difference tells us how much deaths fell due to natural subsidence. Subtracting the two leaves only the pump removal effect.

### Why DiD Works: The Parallel Trends Assumption

DiD relies on one key assumption: **parallel trends**. Without pump removal, both areas would have experienced the same epidemic trajectory.

What this means:

- Areas don't need the same baseline death rates
- They just need to follow the same epidemic curve (same slope)
- Any divergence after pump removal is attributed to the intervention

## 3. John Snow's Cholera Study: A Natural DiD Design

In 1854, London experienced a devastating cholera outbreak. Dr. John Snow famously traced the outbreak to contaminated water from the Broad Street pump. On September 8, 1854, the pump handle was removed, creating a natural experiment.

We'll use the `HistData` package's Snow dataset, which contains weekly cholera deaths. We'll construct a DiD analysis comparing:

- **Treatment group**: Areas primarily served by Broad Street pump (high exposure)
- **Control group**: Areas served by other water sources (low exposure)
- **Treatment**: Pump handle removal (September 8, 1854)

### Data Preparation

```{r prepare-cholera-data}
data(Snow.deaths)
data(Snow.streets)
data(Snow.pumps)

# Create weekly data from late August through late September
weeks <- seq(as.Date("1854-08-21"), as.Date("1854-09-25"), by = "week")
week_labels <- c("Week -2", "Week -1", "Week 0", "Week 1", "Week 2", "Week 3")

# Define districts by proximity to Broad Street pump (pump 4 in the data)
broad_st_pump <- Snow.pumps[4, ]

# Calculate distances from each death to Broad Street pump
Snow.deaths$dist_to_broad <- sqrt(
  (Snow.deaths$x - broad_st_pump$x)^2 + 
  (Snow.deaths$y - broad_st_pump$y)^2
)

# Create treatment assignment: close to Broad Street = treatment
median_dist <- median(Snow.deaths$dist_to_broad)
Snow.deaths$treat <- ifelse(Snow.deaths$dist_to_broad < median_dist, 1, 0)

# Simulating weekly pattern with perfect parallel trends
set.seed(123)

# Create weekly panel
cholera_data <- expand.grid(
  week = 0:5,
  district = 1:40
) %>%
  mutate(
    treat = rep(c(rep(1, 20), rep(0, 20)), 6),
    post = ifelse(week >= 2, 1, 0),  # Week 2 = pump removal
    week_date = weeks[week + 1],
    week_label = week_labels[week + 1]
  )

# Simulate death counts with DiD structure - PERFECT PARALLEL TRENDS
cholera_data <- cholera_data %>%
  mutate(
    # Baseline: treatment areas have higher baseline risk (time-invariant)
    baseline = ifelse(treat == 1, 12, 8),
    
    # Common epidemic trend (IDENTICAL for both groups)
    epidemic_trend = case_when(
      week == 0 ~ 5,
      week == 1 ~ 15,
      week == 2 ~ 10,
      week == 3 ~ 4,
      week == 4 ~ 2,
      week == 5 ~ 1
    ),
    
    # Treatment effect: pump removal reduces deaths ONLY post-treatment
    treatment_effect = ifelse(treat == 1 & post == 1, -8, 0),
    
    # Deaths with minimal noise for clean parallel trends
    deaths = pmax(0, baseline + epidemic_trend + treatment_effect + 
                    rnorm(n(), 0, 0.5)),  # Reduced noise from 1 to 0.5
    
    # Group labels
    group = ifelse(treat == 1, "Near Broad St Pump", "Other Areas"),
    treat_post = treat * post
  )

# Summary statistics
cholera_data %>%
  group_by(group, week_label) %>%
  summarize(
    n_districts = n_distinct(district),
    total_deaths = sum(deaths),
    mean_deaths = mean(deaths),
    sd_deaths = sd(deaths),
    .groups = "drop"
  ) %>%
  kable(digits = 2, caption = "Weekly Cholera Deaths by Area and Week") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## 4. Visualizing the DiD Design

```{r cholera-did-visualization, fig.width=10, fig.height=8}
# Calculate group means by week
group_means <- cholera_data %>%
  group_by(group, week, week_label) %>%
  summarize(mean_deaths = mean(deaths), .groups = "drop")

# Counterfactual: treatment group's pre-treatment path + control group's post-treatment trend
treat_pre <- group_means %>% 
  filter(group == "Near Broad St Pump", week <= 1)

treat_week1 <- group_means %>% 
  filter(group == "Near Broad St Pump", week == 1)
control_week1 <- group_means %>% 
  filter(group == "Other Areas", week == 1)
control_post <- group_means %>% 
  filter(group == "Other Areas", week >= 2)

counterfactual <- bind_rows(
  treat_pre %>% mutate(group = "Counterfactual"),
  control_post %>%
    mutate(
      group = "Counterfactual",
      mean_deaths = treat_week1$mean_deaths + 
        (mean_deaths - control_week1$mean_deaths)
    )
)

plot_data <- bind_rows(group_means, counterfactual)

ggplot() +
  geom_line(data = cholera_data, 
            aes(x = week, y = deaths, group = district),
            color = "grey70", alpha = 0.3, linewidth = 0.5) +
  geom_line(data = filter(plot_data, group %in% c("Near Broad St Pump", "Other Areas")),
            aes(x = week, y = mean_deaths, color = group), linewidth = 1.5) +
  geom_line(data = filter(plot_data, group == "Counterfactual"),
            aes(x = week, y = mean_deaths, color = "Counterfactual"),
            linetype = "dashed", linewidth = 1.5) +
  geom_point(data = filter(plot_data, group %in% c("Near Broad St Pump", "Other Areas")),
             aes(x = week, y = mean_deaths, color = group), size = 3) +
  geom_point(data = filter(plot_data, group == "Counterfactual"),
             aes(x = week, y = mean_deaths, color = "Counterfactual"), size = 2) +
  geom_vline(xintercept = 1.5, linetype = "dashed", alpha = 0.5, color = "gray25") +
  annotate("text", x = 1.5, y = max(plot_data$mean_deaths),
           label = "Pump Removal\n(Sept 8, 1854)", vjust = -0.5, size = 4) +
  # DiD Effect bracket
  annotate("segment",
           x = 5.1, xend = 5.1,
           y = filter(plot_data, group == "Near Broad St Pump", week == 5)$mean_deaths,
           yend = filter(plot_data, group == "Counterfactual", week == 5)$mean_deaths,
           color = "black", linewidth = 1) +
  annotate("segment",
           x = 5.05, xend = 5.1,
           y = filter(plot_data, group == "Near Broad St Pump", week == 5)$mean_deaths,
           yend = filter(plot_data, group == "Near Broad St Pump", week == 5)$mean_deaths,
           color = "black", linewidth = 1) +
  annotate("segment",
           x = 5.05, xend = 5.1,
           y = filter(plot_data, group == "Counterfactual", week == 5)$mean_deaths,
           yend = filter(plot_data, group == "Counterfactual", week == 5)$mean_deaths,
           color = "black", linewidth = 1) +
  annotate("text", x = 5.25,
           y = mean(c(
             filter(plot_data, group == "Near Broad St Pump", week == 5)$mean_deaths,
             filter(plot_data, group == "Counterfactual", week == 5)$mean_deaths
           )),
           label = "DiD Effect >>", hjust = 1.3, vjust = -2, size = 5, fontface = "bold") +
  scale_color_manual(values = c(
    "Near Broad St Pump" = "red3",
    "Other Areas" = "blue3",
    "Counterfactual" = "pink"
  )) +
  scale_x_continuous(breaks = 0:5, labels = week_labels) +
  labs(
    title = "Cholera Deaths Before and After Broad Street Pump Removal",
    subtitle = "Grey lines = individual districts; bold lines = group averages",
    x = "Week (Pump Removed in Week 2)",
    y = "Weekly Deaths",
    color = "Group"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
```

**Key observations:**

1. **Parallel pre-trends (Weeks -2 to 1)**: Both areas follow similar epidemic trajectory
2. **Treatment divergence (Week 2+)**: Areas near Broad Street show sharper decline after pump removal
3. **Counterfactual**: Dashed line shows where Broad Street areas would be without intervention
4. **DiD estimate**: Vertical gap shows lives saved by pump removal

## 5. The DiD Calculation: Step by Step

Using Week 1 (pre-removal) and Week 5 (post-removal) for classic 2×2 comparison:

```{r manual-cholera-did}
# Filter to 2×2 design: Week 1 vs Week 5
cholera_2x2 <- cholera_data %>%
  filter(week %in% c(1, 5)) %>%
  mutate(period = ifelse(week == 1, "Pre (Week 1)", "Post (Week 5)"))

# Calculate means
means_table <- cholera_2x2 %>%
  group_by(group, period) %>%
  summarize(mean_deaths = mean(deaths), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = mean_deaths)

print(means_table)

# Extract means
treat_post <- means_table %>% 
  filter(group == "Near Broad St Pump") %>% 
  pull(`Post (Week 5)`)
treat_pre <- means_table %>% 
  filter(group == "Near Broad St Pump") %>% 
  pull(`Pre (Week 1)`)
control_post <- means_table %>% 
  filter(group == "Other Areas") %>% 
  pull(`Post (Week 5)`)
control_pre <- means_table %>% 
  filter(group == "Other Areas") %>% 
  pull(`Pre (Week 1)`)

# Calculate differences
treat_diff <- treat_post - treat_pre
control_diff <- control_post - control_pre
did_estimate <- treat_diff - control_diff

# Display calculation
data.frame(
  Step = c(
    "Broad St areas change (Week 5 - Week 1)",
    "Other areas change (Week 5 - Week 1)", 
    "DiD estimate (Δ_treat - Δ_control)",
    "True effect (simulated)"
  ),
  Calculation = c(
    sprintf("%.2f - %.2f", treat_post, treat_pre),
    sprintf("%.2f - %.2f", control_post, control_pre),
    sprintf("%.2f - (%.2f)", treat_diff, control_diff),
    "-8.00"
  ),
  Result = c(treat_diff, control_diff, did_estimate, -8.00)
) %>%
  kable(digits = 2, caption = "Manual DiD Calculation: Pump Removal Effect") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

**Understanding the calculation:**

- **Broad Street change**: Natural epidemic decline + pump removal effect
- **Other areas change**: Natural epidemic decline only
- **DiD**: Isolates pump removal effect by differencing out common epidemic trend

## 6. Estimating DiD with Regression

```{r cholera-did-regression}
# Basic DiD regression (Week 1 vs Week 5)
cholera_did_model <- feols(deaths ~ treat + post + treat_post, 
                           data = cholera_2x2)

etable(cholera_did_model, digits = 3, se.below = TRUE)
```

**Coefficients:**

- **treat**: Baseline difference (Broad St areas had higher baseline risk)
- **post**: Natural epidemic decline affecting both groups
- **treat_post**: DiD estimator (pump removal effect = lives saved)

### Relationship Between DiD and TWFE

Before estimating the panel model, it helps to understand how DiD and Two-Way Fixed Effects (TWFE) relate to each other.

**DiD** is a **research design** comparing areas near vs. far from the pump before vs. after removal.

**TWFE** is an **estimation method** that controls for district-specific and week-specific factors using fixed effects.

**Key connections:**

1. **They give the same answer** in simple 2×2 cases:
   - DiD: $Y_{it} = \beta_0 + \beta_1 \text{Treat}_i + \beta_2 \text{Post}_t + \beta_3 (\text{Treat}_i \times \text{Post}_t) + \epsilon_{it}$
   - TWFE: $Y_{it} = \beta_3 (\text{Treat}_i \times \text{Post}_t) + \alpha_i + \gamma_t + \epsilon_{it}$

   > **Intuition:** In the DiD model, you explicitly include dummy variables for treatment group and post period. In TWFE, district fixed effects ($\alpha_i$) do the same job as the `treat` dummy—they absorb any time-stable difference between districts. Week fixed effects ($\gamma_t$) replace the `post` dummy, absorbing the common epidemic curve. The interaction coefficient $\beta_3$ is identical in both.

2. **TWFE extends DiD** to handle multiple weeks and districts efficiently.

3. **Be careful with staggered treatment**: If pump removals happened at different times across London, TWFE can give biased estimates. Modern DiD methods (Callaway & Sant'Anna, Sun & Abraham) address this issue. Click [here](https://idlhy0218.github.io/page%20building/blog.html#statsLab__staggered_did) for these new methods.

### DiD with Fixed Effects

```{r cholera-did-fe}
# Using full panel (Weeks 0-5) with fixed effects
cholera_fe_model <- feols(deaths ~ treat_post | district + week,
                          data = cholera_data)

etable(cholera_did_model, cholera_fe_model,
       headers = c("2×2 DiD", "TWFE"),
       digits = 3, se.below = TRUE)
```

**Comparing specifications:**

- **Model 1 (2×2 DiD)**: Uses only Week 1 (pre-removal) and Week 5 (post-removal), giving 80 observations. Estimates baseline difference between areas near vs. far from Broad Street (`treat` = 4.20 deaths), natural epidemic decline (`post` = -14.2 deaths), and pump removal effect (`treat_post` = -7.80 deaths saved).

- **Model 2 (TWFE)**: Uses all 6 weeks (240 observations). District fixed effects absorb time-invariant baseline differences between areas. Week fixed effects absorb the natural epidemic curve affecting all districts equally.

**Key differences:**

1. **Precision**: The 2×2 DiD is more efficient (SE = 0.221 vs. 0.458), as TWFE uses clustered standard errors by district, which are generally larger but account for within-district correlation across weeks.
2. **Standard errors**: TWFE clusters by district, accounting for repeated observations within the same district over time, which inflates SE relative to the IID assumption in the 2×2 DiD.
3. **Fit**: The 2×2 DiD shows higher overall R² (0.997 vs. 0.982), but this comparison is somewhat misleading — the TWFE Within R² of 0.772 reflects how well the model explains variation after removing district and week fixed effects, which is the more meaningful benchmark.
4. **Treatment effect**: Both models show significant mortality reduction (−7.80 vs. −6.26 deaths saved), with the 2×2 DiD relying on only two time points while TWFE uses the full six-week trajectory, which likely accounts for the modest difference in magnitude.

   > **Intuition:** TWFE is generally the preferred specification among researchers. By including district and week fixed effects, it accounts for time-invariant differences across districts and common time shocks simultaneously, making the treatment effect estimate more credible.

## 7. Testing Parallel Trends

```{r cholera-parallel-trends, fig.width=10, fig.height=6}
# Group means by week
trends_data <- cholera_data %>%
  group_by(group, week, week_label) %>%
  summarize(mean_deaths = mean(deaths), .groups = "drop")

# Pre-treatment regression by group
pre_data <- trends_data %>% filter(week <= 1)

treat_pre <- pre_data %>% filter(group == "Near Broad St Pump")
control_pre <- pre_data %>% filter(group == "Other Areas")

# Fit pre-trend lines
treat_lm <- lm(mean_deaths ~ week, data = treat_pre)
control_lm <- lm(mean_deaths ~ week, data = control_pre)

# Extract slopes
treat_slope <- coef(treat_lm)[2]
control_slope <- coef(control_lm)[2]

ggplot(trends_data, aes(x = week, y = mean_deaths, color = group)) +
  geom_line(linewidth = 1.5, alpha = 0.4) +
  geom_point(size = 3) +
  # Highlight pre-period with thicker lines
  geom_line(data = pre_data, linewidth = 2) +
  geom_point(data = pre_data, size = 4) +
  # Extended trend lines
  geom_abline(intercept = coef(treat_lm)[1], slope = treat_slope,
              color = "red3", linetype = "dashed", linewidth = 1) +
  geom_abline(intercept = coef(control_lm)[1], slope = control_slope,
              color = "blue3", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 1.5, linetype = "dotted", color = "black") +
  annotate("text", x = 1.5, y = min(trends_data$mean_deaths),
           label = "Pump\nRemoval", vjust = -0.5, size = 3.5, fontface = "bold") +
  # Add slope annotations
  annotate("text", x = 0.5, y = 18.5,
           label = sprintf("Slope = %.2f", treat_slope),
           color = "red3", size = 3) +
  annotate("text", x = 0.5, y = 13.5,
           label = sprintf("Slope = %.2f", control_slope),
           color = "blue3", size = 3) +
  scale_color_manual(values = c(
    "Near Broad St Pump" = "red3",
    "Other Areas" = "blue3"
  )) +
  scale_x_continuous(breaks = 0:5, labels = week_labels) +
  labs(
    title = "Parallel Trends Test: Pre-Removal Period",
    subtitle = "Both areas show similar epidemic growth rates before intervention",
    x = "Week",
    y = "Weekly Deaths",
    color = "Area"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
```

**Testing the parallel trends assumption:**

The parallel trends assumption requires both areas to follow similar trajectories before pump removal. We test this by:

1. **Visual inspection**: Plotting pre-removal trends (Weeks 0-1) for both areas
2. **Slope comparison**: Calculating growth rates in each area before intervention
3. **Extended projections**: Dashed lines show where trends would continue without treatment

**What to look for:**

- Similar slopes in pre-period (indicating parallel epidemic growth)
- No systematic divergence before Week 2
- Post-removal deviation only in treated area

If slopes differ substantially or groups already diverge pre-treatment, the parallel trends assumption fails and DiD estimates may be biased.

## 8. Practical Recommendations

**For credible DiD:**

1. Visualize trends before estimating
2. Test parallel trends with pre-periods
3. Report event studies showing dynamics
4. Cluster standard errors by treatment unit
5. Be transparent about assumption violations

**Common pitfalls:**

- Not testing parallel trends
- Ignoring heterogeneous treatment effects in staggered designs
- Small sample sizes
- Using post-treatment data to "test" parallel trends

**When DiD works best:**

- Clear treatment timing
- Plausible parallel trends (similar units)
- Multiple pre-periods for testing
- Exogenous treatment timing
- Adequate sample size

## 9. Conclusion

Difference-in-differences remains one of the most powerful and intuitive methods for causal inference with observational data. Its core logic—comparing changes over time between treatment and control groups—provides a transparent framework for policy evaluation.

The cholera example illustrates DiD's strength: even without randomization, comparing epidemic trajectories between areas near and far from the contaminated pump isolates the pump removal effect by differencing out the natural epidemic decline.

However, DiD is not a magic bullet. The parallel trends assumption, while weaker than assuming no selection bias, is still a strong identifying assumption that requires careful justification. Modern developments in DiD methodology (particularly for staggered treatments) continue to refine and strengthen the approach. Click [here](https://idlhy0218.github.io/page%20building/blog.html#statsLab__staggered_did) for these new methods!
