---
title: "Machine Learning for Social Scientists, Part 1: Supervised ML - Ridge, Lasso, and Elastic Net"
author: "Heeyoung Lee"
date: "August 12, 2026"
format: 
  html:
    theme: cosmo
    css: custom.css
    page-layout: full
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 5
    toc-location: left
    self-contained: true
    fig-width: 9
    fig-height: 6
    fig-dpi: 300
    smooth-scroll: true
    highlight-style: github
---

## Machine Learning Foundations, Part 1

Every quantitative researcher eventually asks the same question: "Should I trust the coefficient, or should I trust the prediction?" Classical statistics and machine learning give different answers, and understanding *why* they differ is the single most useful thing you can learn before touching a machine learning method. This post covers four foundational ideas — supervised vs. unsupervised learning, two cultures of statistical analysis, the bias-variance tradeoff, and the logic of regularization — using the same county-level health framing from the panel data post, so the intuition carries over.

Let's begin by loading the packages we'll need:

```{r library-imports, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(glmnet)     # For lasso and ridge regression
library(tidyr)
```

### 1. Two Basic Flavors: Supervised vs. Unsupervised Learning

Before we get into cultures and penalties, it helps to sort machine learning methods into two big buckets. Almost everything you'll hear about — lasso, ridge, decision trees, random forests, clustering, topic models — falls into one of these two categories, and knowing which bucket you're in tells you what kind of question you can even ask.

**The one-sentence version: does your data have an "answer key" or not?**

- **Supervised learning** — you have a known outcome you're trying to predict. Every county in your data has both the predictors *and* the actual mortality rate attached to it. The algorithm learns by comparing its guesses to the real answer, the way a student learns from a teacher who grades their homework.
- **Unsupervised learning** — there is no known outcome. You just have a pile of county characteristics, and you're asking the algorithm to find structure or patterns in it on its own — no teacher, no answer key, no grading.

#### A Health Example of Each

Let's stick with our running county mortality example and see how the same dataset supports both kinds of tasks.

**Supervised task**: "Using each county's poverty rate, uninsured rate, obesity rate, and physician density, predict its mortality rate." Here, mortality rate is the known outcome (the "label") sitting right there in the data for every county we've already observed. The whole point of the exercise is to get better and better at guessing it. This is exactly the kind of task that lasso, ridge, decision trees, and random forests are built for — everything covered in the rest of this post, and in most of what people mean by "machine learning" in applied research.

**Unsupervised task**: "Group the 3,000 U.S. counties into a handful of health profiles, based on their poverty rate, obesity rate, opioid prescribing rate, and rural/urban status — without telling the algorithm what the groups should look like." There's no "correct" grouping to check against. The algorithm (e.g., cluster analysis) just looks for counties that resemble each other and sorts them into groups. You might end up with something like "high-poverty rural counties with elevated opioid prescribing," "urban counties with high obesity but strong healthcare access," and so on — categories nobody specified in advance, but that emerged from the data itself.

```{r supervised-vs-unsupervised, fig.width=11, fig.height=5}
set.seed(88)
n_demo <- 80

# Simulate two underlying "true" county health profiles for illustration
profile <- sample(c("A", "B"), n_demo, replace = TRUE)
poverty_demo <- ifelse(profile == "A", rnorm(n_demo, 22, 4), rnorm(n_demo, 10, 4))
obesity_demo <- ifelse(profile == "A", rnorm(n_demo, 34, 4), rnorm(n_demo, 26, 4))
mortality_demo <- 400 + 5 * poverty_demo + 3 * obesity_demo + rnorm(n_demo, 0, 15)

demo_df <- data.frame(poverty = poverty_demo, obesity = obesity_demo, 
                       mortality = mortality_demo, profile = profile)

# Supervised view: predictors colored by the known outcome (mortality)
p_supervised <- ggplot(demo_df, aes(x = poverty, y = obesity, color = mortality)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_gradient(low = "gold", high = "firebrick", name = "Mortality\nRate") +
  labs(title = "Supervised Learning",
       subtitle = "Mortality rate is known\nfor every county",
       x = "Poverty Rate (%)", y = "Obesity Rate (%)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.margin = margin(10, 15, 10, 10))

# Unsupervised view: same predictors, no outcome shown, just raw points
p_unsupervised <- ggplot(demo_df, aes(x = poverty, y = obesity)) +
  geom_point(size = 3, alpha = 0.8, color = "gray40") +
  labs(title = "Unsupervised Learning",
       subtitle = "No known outcome —\nonly patterns among counties",
       x = "Poverty Rate (%)", y = "Obesity Rate (%)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.margin = margin(10, 15, 10, 10))

p_supervised + p_unsupervised
```

Look at the two panels above. They show the *exact same* 80 counties, plotted on the same two axes (poverty rate and obesity rate). The only difference is what information we're allowed to use:

- **Left (supervised)**: each county is colored by its actual mortality rate (gold = low, red = high). The algorithm's job is to learn the relationship between the x/y position and that color, so it can predict the color (mortality rate) for a brand-new county.
- **Right (unsupervised)**: the color is hidden — we've stripped away the outcome entirely. All the algorithm can do is notice that the points seem to fall into two loose clusters based on their position alone. It has no way of knowing whether those clusters mean anything about mortality; it's just describing structure in the predictors themselves.

#### Why This Distinction Matters Before You Choose a Method

Every method we'll walk through in the rest of this post — and the next one — sits cleanly in one of these two buckets:

| | Supervised | Unsupervised |
|---|---|---|
| **Has a "right answer" during training?** | Yes | No |
| **Typical question** | "Predict Y from X" | "Find hidden structure in X" |
| **Examples** | Lasso, ridge, elastic net, decision trees, random forests, logistic regression, k-NN | Cluster analysis, principal component analysis, topic modeling |
| **How you check if it worked** | Compare predictions to the true, known outcome | Judge whether the discovered groups make substantive or statistical sense (no ground truth to check against) |

The rest of this post — two cultures, bias-variance, and regularization — all lives inside the **supervised** bucket, because lasso, ridge, and elastic net are all methods for predicting a known outcome (mortality rate) from a set of predictors. Unsupervised methods raise a different set of questions entirely — mainly, "how do we know the patterns we found are real, and not just noise?" — which we'll return to in a later post.

### 2. Two Cultures of Statistical Analysis

#### The Core Distinction

In 2001, the statistician Leo Breiman wrote a famous paper describing two very different ways that researchers approach data. He called them the "two cultures," and the distinction still organizes how we think about classical statistics versus machine learning today.

**Culture 1 — Generative modeling (classical statistics).** You assume the data were generated by some underlying stochastic process, propose a model for that process (usually linear), and estimate its parameters. The goal is **inference**: understanding *how* an output relates to inputs.

**Culture 2 — Predictive modeling (machine learning).** You treat the true data-generating process as unknown and unknowable. Instead of trying to describe it, you search for a function that predicts the output as accurately as possible on *new* data. The goal is **prediction**, not explanation.

This is not just a philosophical distinction — it changes what a "good model" even means.

| | Classical Statistics (Culture 1) | Machine Learning (Culture 2) |
|---|---|---|
| **Goal** | Estimate $\hat{\beta}$ — understand relationships | Estimate $\hat{Y}$ — predict new outcomes |
| **Question asked** | "How much does income affect mortality?" | "Given this county's profile, what's its mortality rate?" |
| **Success measured by** | Coefficient significance, theoretical fit | Accuracy on data the model has never seen |
| **Model complexity** | Kept small and interpretable | Can be very large and complex |
| **Risk if misused** | Missing real complexity in the world | Producing a black box that predicts well but explains nothing |

#### A Concrete Example

Suppose we're back in our county mortality dataset. If we're doing **inference**, our question is: "Holding other things constant, does a $1,000 increase in county median income reduce mortality, and by how much?" We want a trustworthy $\hat{\beta}$.

If we're doing **prediction**, our question is completely different: "A new county just reported its income, uninsured rate, obesity rate, and physician density. What mortality rate should we expect?" We don't care whether income "causes" anything — we just want $\hat{Y}$ to be close to the county's actual mortality rate.

```{r two-cultures-sim, fig.width=9}
set.seed(42)
n <- 300

# Simulate a nonlinear true relationship between income and mortality
income <- runif(n, 30, 90)
true_mortality <- 900 - 6 * income + 0.04 * income^2 + rnorm(n, 0, 25)

sim_df <- data.frame(income = income, mortality = true_mortality)

# Culture 1: a simple linear model (inference-focused)
linear_fit <- lm(mortality ~ income, data = sim_df)

# Culture 2: a flexible model (prediction-focused) - local smoothing
sim_df$loess_pred <- predict(loess(mortality ~ income, data = sim_df, span = 0.5))

ggplot(sim_df, aes(x = income, y = mortality)) +
  geom_point(alpha = 0.4, color = "gray40") +
  geom_smooth(method = "lm", se = FALSE, aes(color = "Linear model\n(inference-style)"), linewidth = 1.2) +
  geom_line(aes(y = loess_pred, color = "Flexible curve\n(prediction-style)"), linewidth = 1.2) +
  labs(title = "Same Data, Two Different Goals",
       subtitle = "A straight line is easy to interpret;\na flexible curve fits the data more closely",
       x = "Median Household Income ($1000s)",
       y = "Mortality Rate (per 100,000)",
       color = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 14),
        plot.margin = margin(10, 15, 10, 10))
```

Both lines are "correct" in their own way. The straight line gives us one clean number — a single $\hat{\beta}$ we can put in a sentence: "each additional $1,000 in income is associated with a X-unit change in mortality." The curved line gives up that clean sentence, but it tracks the *actual* pattern in the data more closely, which matters if our only goal is predicting mortality for a new county.

**The key takeaway:** neither approach is "better" in general. The right choice depends on whether your research question is "why" (inference) or "what will happen" (prediction).

### 3. The Bias-Variance Tradeoff and Generalization Error

#### Why a Model That Fits Well Can Still Predict Badly

Here's a puzzle that trips up a lot of researchers the first time they see it: a model can fit your existing data almost perfectly and still make *terrible* predictions on new data. Understanding why requires breaking prediction error into two separate pieces: **bias** and **variance**.

**Bias** is the error from a model that's too simple to capture the real pattern. Think of a clock that always runs an hour late — every reading is wrong in the same, predictable direction. A straight line fit to a curved relationship has bias: it will systematically miss the curve.

**Variance** is the error from a model that's *too* sensitive to the specific data it happened to see. Think of a clock that randomly runs fast or slow — you can't predict which way it will be wrong next time. A model with high variance would look completely different if you gave it a slightly different sample of the same population.

$$\text{Generalization Error} = \text{Bias}^2 + \text{Variance} + \text{Irreducible Error}$$

The irreducible error is just noise in the world that no model can ever remove — measurement error, unpredictable individual circumstances, and so on.

#### Underfitting and Overfitting

- **Underfitting** happens when the model is too simple: it misses real signal in the data. High bias, low variance.
- **Overfitting** happens when the model is too complex: it captures not just the signal, but also the random noise specific to *this* sample. Low bias, high variance.

#### The Dartboard Picture (the Classic Way to See Bias vs. Variance)

Before we bring back the county data, here's a picture that makes bias and variance click for a lot of people. Imagine four different dart players, each throwing 20 darts at the same target. The bullseye represents the *true* mortality rate we're trying to hit; each dart is one prediction from repeating the same model on slightly different samples of counties.

```{r dartboard, fig.width=9, fig.height=9}
make_target <- function(n = 20, center_offset = c(0, 0), spread = 0.15, title = "", subtitle = "") {
  set.seed(1)
  darts <- data.frame(
    x = rnorm(n, mean = center_offset[1], sd = spread),
    y = rnorm(n, mean = center_offset[2], sd = spread)
  )
  
  circle_data <- function(r) {
    theta <- seq(0, 2 * pi, length.out = 100)
    data.frame(x = r * cos(theta), y = r * sin(theta))
  }
  
  ggplot() +
    geom_polygon(data = circle_data(1), aes(x, y), fill = "#f4d1d1", color = "gray60") +
    geom_polygon(data = circle_data(0.66), aes(x, y), fill = "#f9e8b0", color = "gray60") +
    geom_polygon(data = circle_data(0.33), aes(x, y), fill = "#c9e8c9", color = "gray60") +
    geom_point(data = darts, aes(x, y), size = 3, color = "black", alpha = 0.75) +
    coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3)) +
    labs(title = title, subtitle = subtitle) +
    theme_void(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
          plot.subtitle = element_text(size = 10, hjust = 0.5),
          plot.margin = margin(8, 8, 8, 8))
}

p1 <- make_target(center_offset = c(0, 0), spread = 0.10, 
                   title = "Low Bias, Low Variance", subtitle = "Ideal: accurate and consistent")
p2 <- make_target(center_offset = c(0, 0), spread = 0.35, 
                   title = "Low Bias, High Variance", subtitle = "Overfitting: centered, but scattered")
p3 <- make_target(center_offset = c(0.55, 0.45), spread = 0.10, 
                   title = "High Bias, Low Variance", subtitle = "Underfitting: consistent, but off-target")
p4 <- make_target(center_offset = c(0.55, 0.45), spread = 0.35, 
                   title = "High Bias, High Variance", subtitle = "Worst case: inaccurate and inconsistent")

(p1 + p2) / (p3 + p4)
```

Read this like a real dartboard: the bullseye (green center) is the truth. Each of the 20 dots is what the model would have predicted if it had been trained on a slightly different sample of counties.

- **Top-left**: darts cluster tightly around the bullseye — this is what we want. Low bias (correct on average), low variance (consistent across samples).
- **Top-right**: darts are centered on the bullseye *on average*, but wildly scattered — this is **overfitting**. Any one run might land far from the truth, even though the model isn't systematically wrong in one direction.
- **Bottom-left**: darts cluster tightly, but off to one side — this is **underfitting**. The model is consistent, but consistently *wrong* the same way every time.
- **Bottom-right**: darts are both off-target and scattered — the worst of both problems at once.

Now let's make this concrete with our county data. Suppose the true relationship between income and mortality has a slight curve to it (as in the simulation above), but we don't know that — we only have a sample of counties to learn from.

```{r bias-variance-sim, fig.height=8}
set.seed(7)

# True underlying function (unknown to the researcher in practice)
true_function <- function(x) 900 - 6 * x + 0.04 * x^2

# Generate one sample
n_sample <- 40
x_sample <- runif(n_sample, 30, 90)
y_sample <- true_function(x_sample) + rnorm(n_sample, 0, 25)
sample_df <- data.frame(income = x_sample, mortality = y_sample)

x_grid <- seq(30, 90, length.out = 200)
true_line <- data.frame(income = x_grid, mortality = true_function(x_grid))

# Underfit: a flat model (intercept only)
underfit_pred <- data.frame(income = x_grid, mortality = mean(y_sample))

# Good fit: quadratic model (matches the true functional form)
good_fit <- lm(mortality ~ poly(income, 2), data = sample_df)
goodfit_pred <- data.frame(income = x_grid, 
                            mortality = predict(good_fit, newdata = data.frame(income = x_grid)))

# Overfit: a very high-degree polynomial that chases every point
overfit_model <- lm(mortality ~ poly(income, 15), data = sample_df)
overfit_pred <- data.frame(income = x_grid, 
                            mortality = predict(overfit_model, newdata = data.frame(income = x_grid)))

p_under <- ggplot() +
  geom_point(data = sample_df, aes(income, mortality), alpha = 0.6, color = "gray30") +
  geom_line(data = true_line, aes(income, mortality), color = "black", linetype = "dashed") +
  geom_line(data = underfit_pred, aes(income, mortality), color = "steelblue", linewidth = 1.3) +
  labs(title = "Underfitting", subtitle = "Model too simple — misses the curve (high bias)",
       x = "Income ($1000s)", y = "Mortality Rate") +
  theme_minimal(base_size = 12) +
  theme(plot.margin = margin(8, 15, 8, 8))

p_good <- ggplot() +
  geom_point(data = sample_df, aes(income, mortality), alpha = 0.6, color = "gray30") +
  geom_line(data = true_line, aes(income, mortality), color = "black", linetype = "dashed") +
  geom_line(data = goodfit_pred, aes(income, mortality), color = "darkgreen", linewidth = 1.3) +
  labs(title = "Good Fit", subtitle = "Captures the signal without chasing noise",
       x = "Income ($1000s)", y = "Mortality Rate") +
  theme_minimal(base_size = 12) +
  theme(plot.margin = margin(8, 15, 8, 8))

p_over <- ggplot() +
  geom_point(data = sample_df, aes(income, mortality), alpha = 0.6, color = "gray30") +
  geom_line(data = true_line, aes(income, mortality), color = "black", linetype = "dashed") +
  geom_line(data = overfit_pred, aes(income, mortality), color = "firebrick", linewidth = 1.3) +
  labs(title = "Overfitting", subtitle = "Model too flexible — chases noise (high variance)",
       x = "Income ($1000s)", y = "Mortality Rate") +
  theme_minimal(base_size = 12) +
  theme(plot.margin = margin(8, 15, 8, 8))

p_under / p_good / p_over
```

The dashed black line is the *true* relationship (which we'd never actually know in real research — it's shown here only because this is a simulation). Notice:

- **Underfitting (top)**: the flat blue line ignores the curve entirely. It will be wrong in the same direction for every new county with low or high income.
- **Good fit (middle)**: the green curve tracks the true pattern closely without overreacting to any single point.
- **Overfitting (bottom)**: the red line wiggles wildly to pass as close as possible to every single training point — including the random noise. This model would look completely different if we resampled 40 new counties.

#### Seeing It Numerically: In-Sample vs. Out-of-Sample Error

The real danger of overfitting is invisible if you only look at how well a model fits the data it was trained on. You have to test it on *new* data to see the problem.

```{r train-test-demo}
set.seed(101)

# Generate training data and a separate, fresh test set from the same true process
train_x <- runif(60, 30, 90)
train_y <- true_function(train_x) + rnorm(60, 0, 25)
train_df <- data.frame(income = train_x, mortality = train_y)

test_x <- runif(60, 30, 90)
test_y <- true_function(test_x) + rnorm(60, 0, 25)
test_df <- data.frame(income = test_x, mortality = test_y)

# Fit models of increasing flexibility (polynomial degree = complexity)
degrees <- 1:12
results <- data.frame()

for (d in degrees) {
  model <- lm(mortality ~ poly(income, d), data = train_df)
  train_pred <- predict(model)
  test_pred <- predict(model, newdata = test_df)
  
  train_rmse <- sqrt(mean((train_df$mortality - train_pred)^2))
  test_rmse <- sqrt(mean((test_df$mortality - test_pred)^2))
  
  results <- rbind(results, data.frame(degree = d, 
                                        train_rmse = train_rmse, 
                                        test_rmse = test_rmse))
}

results_long <- results %>%
  pivot_longer(cols = c(train_rmse, test_rmse), names_to = "type", values_to = "rmse") %>%
  mutate(type = ifelse(type == "train_rmse", "Training error (in-sample)", "Test error (out-of-sample)"))

ggplot(results_long, aes(x = degree, y = rmse, color = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(title = "Training Error Keeps Falling — Test Error Does Not",
       subtitle = "Model complexity (polynomial degree) on the x-axis",
       x = "Model Complexity (polynomial degree)",
       y = "Root Mean Squared Error (RMSE)",
       color = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 13),
        plot.margin = margin(10, 15, 10, 10))
```

This is the single most important picture in this post. It shows what happens as we make a model more and more complex (by raising the polynomial degree).

**How the experiment works**: We split the data into two separate groups.

- **Training data**: the data the model actually learns from. For example, we show the model 100 counties and let it learn the pattern from them.

- **Test data**: a separate set of counties the model has never seen before. This is used to check how well the model would perform on a brand-new county in the real world.

We then make the model more and more complex — degree 1, degree 2, degree 3, and so on — and watch how the error (RMSE) changes on each of the two datasets.

**What we observe**:

- **Training error keeps shrinking.** As the model gets more complex, it can always fit the data it has already seen more closely — like memorizing the answers to a test you've already taken, getting closer and closer to a perfect score.

- **Test error falls at first, then turns around and starts rising again.** At first, added complexity helps the model capture more of the real underlying pattern (the "signal"). But past a certain point, the model starts memorizing the random noise mixed into the training data. That noise doesn't repeat in a new county's data, so from that point on, predictions on new data actually get worse.

**That turning point is the key.** Before it, the model is learning real signal. After it, the model has started memorizing noise.

**Bottom line**: the goal of any good model — classical statistics or machine learning — is to land at the point where **test error** is lowest, not where training error is lowest. If you judge a model only by its training error, you can end up picking a model that looks great on paper but performs poorly on any new county it hasn't already memorized.


### 4. The Principle of Regularization

#### The Core Idea

Now we get to the tool that machine learning uses to avoid landing on the wrong side of that curve: **regularization**. The idea is simple to state: instead of only minimizing error on the training data, add a penalty for model complexity.

$$\underbrace{\sum_{i=1}^{n}[y_i - f(x_i)]^2}_{\text{fit to training data}} \;+\; \underbrace{\lambda R(f)}_{\text{penalty for complexity}}$$

- The first term is the same thing OLS already minimizes: how far off are our predictions?
- The second term, $\lambda R(f)$, punishes the model for being too flexible or having too-large coefficients.
- $\lambda$ (lambda) controls *how much* we care about that penalty. In ordinary OLS, $\lambda = 0$ — no penalty at all. As $\lambda$ increases, we push the model toward simplicity.

Choosing $\lambda$ is itself data-driven — it's typically selected via cross-validation, so the model doesn't just optimize the penalty by guesswork.

#### Lasso, Ridge, and Elastic Net: Three Flavors of Penalty

The three primary regularized regression methods differ in how they define the penalty term $R(f)$: All three of these methods start from the exact same idea — penalize big coefficients — but they punish "bigness" in different ways, and that difference changes what the model does in practice.

Think of it like packing a suitcase with a strict weight limit. You have 30 items (our county-level predictors: poverty rate, opioid prescribing rate, racial composition, physician density, and so on) and a budget for how much "weight" your model's coefficients are allowed to carry in total.


#### 1. Lasso ($L_1$ Penalty)

"Some items just don't make the cut — leave them home entirely." Lasso can shrink some coefficients all the way to *exactly* zero, which means those variables are dropped from the model completely. This gives you automatic variable selection: lasso tells you which handful of predictors actually matter.

$$\min_{\beta} \sum_{i=1}^{n} [y_i - f(x_i)]^2 \;+\; \lambda \sum_{j=1}^{p} \lvert\beta_j\rvert$$

- **Mechanism**: Punishes the sum of *absolute* values of coefficients.
- **Behavior**: The $L_1$ penalty retains a constant derivative down to zero, driving coefficients of non-essential variables to **exactly zero**. Performs automatic feature selection.

#### 2. Ridge Regression ($L_2$ Penalty)

"Every item can come, but the heavier ones get squeezed down more." Ridge shrinks *every* coefficient toward zero, proportionally more for large ones, but essentially never sets a coefficient to exactly zero. Every variable stays in the model, just with a smaller voice.

$$\min_{\beta} \sum_{i=1}^{n} [y_i - f(x_i)]^2 \;+\; \lambda \sum_{j=1}^{p} \beta_j^2$$

- **Mechanism**: Punishes the sum of *squared* coefficients.
- **Behavior**: Shrinks all coefficients continuously toward zero, but **never sets any coefficient to exactly zero**. Every variable remains in the model.

#### 3. Elastic Net (Blended Penalty)

Compromise between the Lasso and Ridge: "Squeeze everything down, *and* leave some things home." It combines both penalties, weighted by a parameter $\alpha$ between 0 and 1: $\alpha = 1$ is pure lasso, $\alpha = 0$ is pure ridge, and anything in between is elastic net.

$$\min_{\beta} \sum_{i=1}^{n} [y_i - f(x_i)]^2 \;+\; \lambda \left[ \alpha \sum_{j=1}^{p} \lvert\beta_j\rvert + (1-\alpha) \sum_{j=1}^{p} \beta_j^2 \right]$$

- **Mechanism**: Combines $L_1$ (Lasso) and $L_2$ (Ridge) penalties via weight parameter $\alpha \in [0, 1]$.
- **Behavior**: Provides sparse variable selection (via $L_1$) while maintaining stable coefficient grouping among correlated predictors (via $L_2$).

**Why would you ever want ridge instead of lasso, if lasso does variable selection for you?** Two practical reasons come up constantly in county health research:

1. **Correlated predictors.** County-level variables are often highly correlated with each other — for example, poverty rate, uninsured rate, and low education tend to move together. Lasso tends to arbitrarily pick just *one* of a group of correlated variables and zero out the rest, even if they're all genuinely related to the outcome. Ridge instead spreads the credit across all of them, which can be more stable and realistic.
2. **You don't actually want to drop variables.** Sometimes you have good theoretical reasons to keep every predictor in the model (e.g., you're building a prediction tool that should always report physician density, even if its coefficient is small), and you just want to control overfitting without deleting anything. Ridge does exactly that.

**Elastic net** is often the practical default when you have many correlated county-level variables and also want some of them to drop out: it handles correlated groups better than lasso alone, while still allowing some coefficients to hit exactly zero.

#### A Worked Example: Predicting County Mortality

To compare these methods empirically, we simulate a county health dataset with $n = 100$ counties and $p = 30$ candidate predictors. In the true underlying data-generating process, only the first 4 variables drive mortality ($\beta_1=8, \beta_2=-6, \beta_3=5, \beta_4=-4$), while the remaining 26 variables are pure noise ($\beta_{5\dots30} = 0$).

**1. Lasso in Action ($L_1$ Penalty)**

We first fit Lasso ($\alpha = 1$) across a continuum of penalty values ($\lambda$):

```{r lasso-sim}
set.seed(2024)
n_counties_sim <- 100
n_predictors <- 30  # 30 candidate variables

# Simulate predictors (standardized county-level variables)
X <- matrix(rnorm(n_counties_sim * n_predictors), nrow = n_counties_sim)
colnames(X) <- paste0("var", 1:n_predictors)

# Only the first 4 variables genuinely drive mortality
true_betas <- c(8, -6, 5, -4, rep(0, n_predictors - 4))
mortality_sim <- X %*% true_betas + rnorm(n_counties_sim, 0, 5)

# Fit lasso across a range of lambda values
lasso_fit <- glmnet(X, mortality_sim, alpha = 1)  # alpha = 1 is lasso

# Extract coefficient paths
coef_matrix <- as.matrix(coef(lasso_fit))[-1, ]  # remove intercept
coef_df <- as.data.frame(t(coef_matrix)) %>%
  mutate(lambda = lasso_fit$lambda) %>%
  pivot_longer(-lambda, names_to = "variable", values_to = "coefficient") %>%
  mutate(truly_matters = ifelse(variable %in% paste0("var", 1:4), "Actually matters", "Actually zero"))

ggplot(coef_df, aes(x = log(lambda), y = coefficient, group = variable, color = truly_matters)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Lasso Coefficient Path: Shrinkage and Variable Selection",
       subtitle = "As penalty lambda increases, noise variables (gray) drop to zero first",
       x = expression(log(lambda) ~ "(more penalty" %->% ")"),
       y = "Coefficient Value",
       color = "") +
  scale_color_manual(values = c("Actually matters" = "firebrick", "Actually zero" = "gray70")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 13),
        plot.margin = ggplot2::margin(10, 15, 10, 10))
```

**Lasso Interpretation**:
- At low $\lambda$ ($\log(\lambda)$ on left), all 30 variables have non-zero coefficients (similar to unpenalized OLS).
- As $\lambda$ increases, the 26 noise variables (gray lines) drop to zero one by one.
- The 4 true signals (red lines) resist shrinkage longer, demonstrating how Lasso isolates true drivers through **sparse variable selection**.

**2. Ridge in Action ($L_2$ Penalty)**

Next, we fit Ridge Regression ($\alpha = 0$) on the exact same dataset:

```{r ridge-sim}
# Fit ridge across a range of lambda values
ridge_fit <- glmnet(X, mortality_sim, alpha = 0)  # alpha = 0 is ridge

# Extract coefficient paths
ridge_coef_matrix <- as.matrix(coef(ridge_fit))[-1, ]
ridge_coef_df <- as.data.frame(t(ridge_coef_matrix)) %>%
  mutate(lambda = ridge_fit$lambda) %>%
  pivot_longer(-lambda, names_to = "variable", values_to = "coefficient") %>%
  mutate(truly_matters = ifelse(variable %in% paste0("var", 1:4), "Actually matters", "Actually zero"))

ggplot(ridge_coef_df, aes(x = log(lambda), y = coefficient, group = variable, color = truly_matters)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Ridge Coefficient Path: Continuous Shrinkage Without Selection",
       subtitle = "As penalty lambda increases, all 30 coefficients shrink toward zero but never reach zero",
       x = expression(log(lambda) ~ "(more penalty" %->% ")"),
       y = "Coefficient Value",
       color = "") +
  scale_color_manual(values = c("Actually matters" = "firebrick", "Actually zero" = "gray70")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 13),
        plot.margin = ggplot2::margin(10, 15, 10, 10))
```

**Ridge Interpretation**:
- As $\lambda$ increases, all 30 coefficients shrink smoothly toward zero.
- Crucially, **none of the 26 noise variables hit zero exactly**. Every single variable remains in the model, regardless of how large $\lambda$ becomes.

#### Lasso vs. Ridge, Side by Side

Placing Lasso ($\alpha = 1$) and Ridge ($\alpha = 0$) side by side on the exact same dataset highlights their fundamental structural difference:

```{r lasso-vs-ridge, fig.width=11}
lasso_coef_df_labeled <- coef_df %>%
  mutate(method = "Lasso: sparse selection (zeroes noise)")

ridge_coef_df_labeled <- ridge_coef_df %>%
  mutate(method = "Ridge: continuous shrinkage (keeps all 30)")

combined_df <- bind_rows(lasso_coef_df_labeled, ridge_coef_df_labeled)
combined_df$method <- factor(combined_df$method, levels = c("Lasso: sparse selection (zeroes noise)", "Ridge: continuous shrinkage (keeps all 30)"))

ggplot(combined_df, aes(x = log(lambda), y = coefficient, group = variable, color = truly_matters)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~method) +
  labs(title = "Lasso vs. Ridge Coefficient Paths on the Same Data",
       x = expression(log(lambda) ~ "(more penalty" %->% ")"),
       y = "Coefficient Value",
       color = "") +
  scale_color_manual(values = c("Actually matters" = "firebrick", "Actually zero" = "gray70")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 13),
        strip.text = element_text(face = "bold", size = 10.5),
        plot.margin = ggplot2::margin(10, 15, 10, 10))
```

**Key Differences Illustrated**:
- **Lasso (Left)**: Noise lines collapse directly onto the zero axis, performing **hard variable elimination** (setting 26 noise variables to exactly 0).
- **Ridge (Right)**: All 26 gray noise lines shrink continuously toward zero, but **none reach zero**. All 30 variables remain in the model.

#### Choosing the Right Amount of Penalty: Cross-Validation

How do we choose $\lambda$? We don't guess — we use $k$-fold cross-validation (`cv.glmnet()`) to find the penalty that minimizes out-of-sample prediction error. The data is split into $k$ folds; the model trains on $k-1$ folds and tests on the remaining fold, repeating until every fold serves as the test set once.

```{r cv-lambda-lasso, fig.height=6}
set.seed(2024)
cv_lasso <- cv.glmnet(X, mortality_sim, alpha = 1)

par(mar = c(7, 4, 4, 2))
plot(cv_lasso)
title("Lasso: Cross-Validated Error by Lambda", line = 2.5)
mtext("Left dashed line: best prediction (lambda.min)", side = 1, line = 5, cex = 0.8)
mtext("Right dashed line: simplest good model (lambda.1se)", side = 1, line = 6, cex = 0.8)
```

```{r cv-lambda-ridge, fig.height=6}
set.seed(2024)
cv_ridge <- cv.glmnet(X, mortality_sim, alpha = 0)

par(mar = c(7, 4, 4, 2))
plot(cv_ridge)
title("Ridge: Cross-Validated Error by Lambda", line = 2.5)
mtext("Left dashed line: best prediction (lambda.min)", side = 1, line = 5, cex = 0.8)
mtext("Right dashed line: simplest good model (lambda.1se)", side = 1, line = 6, cex = 0.8)
```

**Reading the CV curves**:

- The x-axis is $-\log(\lambda)$ (moving right reduces penalty toward OLS). The y-axis shows mean squared error on held-out folds.
- **Lasso** forms a classic U-shape: error drops as signal is captured, then rises as noise re-enters.
- **Ridge** error decreases continuously across this range as all 30 variables remain active throughout.

```{r cv-lambda-values}
cat("Lasso  — lambda.min:", round(cv_lasso$lambda.min, 3), 
    " | lambda.1se:", round(cv_lasso$lambda.1se, 3), "\n")
cat("Ridge  — lambda.min:", round(cv_ridge$lambda.min, 3), 
    " | lambda.1se:", round(cv_ridge$lambda.1se, 3))
```

We evaluate two standard $\lambda$ thresholds: `lambda.min` (minimum CV error) and `lambda.1se` (the most parsimonious model within one standard error of the minimum).

```{r cv-selected-vars}
# Which variables survive at the cross-validation-selected lambda?
selected_coefs <- coef(cv_lasso, s = "lambda.min")
selected_df <- data.frame(
  variable = rownames(selected_coefs)[-1],
  coefficient = as.vector(selected_coefs)[-1]
) %>%
  filter(coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

kable(selected_df, digits = 2, 
      caption = "Variables Selected by Lasso at the Cross-Validation-Chosen Lambda") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

At `lambda.min`, Lasso retains 12 of the 30 variables. Crucially, the 4 true drivers of mortality (`var1`–`var4`) emerge with by far the largest coefficients, while only 8 minor noise terms slip through.

In contrast, Ridge keeps all 30 variables in the model at its optimal `lambda.min`:

```{r cv-ridge-coefs}
# Ridge coefficients at its own cross-validation-chosen lambda
ridge_selected_coefs <- coef(cv_ridge, s = "lambda.min")
ridge_selected_df <- data.frame(
  variable = rownames(ridge_selected_coefs)[-1],
  coefficient = as.vector(ridge_selected_coefs)[-1]
) %>%
  arrange(desc(abs(coefficient)))

kable(head(ridge_selected_df, 30), digits = 3, 
      caption = "Ridge Coefficients at Its Cross-Validation-Chosen Lambda (Top 8 by Size)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Ridge correctly ranks the 4 true predictors at the top, but shrinks all 30 coefficients continuously rather than performing variable selection.

#### Elastic Net on the Same Data: Combining Ridge and Lasso

How does **Elastic Net** ($\alpha = 0.5$) differ from Lasso ($\alpha = 1$) and Ridge ($\alpha = 0$)? 

Elastic Net blends both penalties: $\lambda [ \alpha \sum |\beta_j| + (1-\alpha) \sum \beta_j^2 ]$. This design addresses the core limitations of both methods:

1. **Unlike Ridge**, Elastic Net can set coefficients to *exactly zero*, dropping noise variables.
2. **Unlike Lasso**, Elastic Net handles *correlated predictors* stably without arbitrarily picking one and zeroing out the rest.

Let's test Elastic Net on the exact same dataset (`X` and `mortality_sim`):

```{r cv-lambda-enet, fig.height=6}
set.seed(2024)
cv_enet <- cv.glmnet(X, mortality_sim, alpha = 0.5)

par(mar = c(7, 4, 4, 2))
plot(cv_enet)
title("Elastic Net (alpha = 0.5): Cross-Validated Error by Lambda", line = 2.5)
mtext("Left dashed line: best prediction (lambda.min)", side = 1, line = 5, cex = 0.8)
mtext("Right dashed line: simplest good model (lambda.1se)", side = 1, line = 6, cex = 0.8)
```

```{r cv-lambda-values-enet}
cat("Ridge        — lambda.min:", round(cv_ridge$lambda.min, 3), 
    " | lambda.1se:", round(cv_ridge$lambda.1se, 3), "\n")
cat("Lasso        — lambda.min:", round(cv_lasso$lambda.min, 3), 
    " | lambda.1se:", round(cv_lasso$lambda.1se, 3), "\n")
cat("Elastic Net  — lambda.min:", round(cv_enet$lambda.min, 3), 
    " | lambda.1se:", round(cv_enet$lambda.1se, 3))
```

Elastic Net's optimal penalty (`lambda.min` = 0.676) falls directly between Ridge (0.799) and Lasso (0.538). 

```{r cv-enet-coefs}
# Which variables survive at elastic net's cross-validation-selected lambda?
enet_selected_coefs <- coef(cv_enet, s = "lambda.min")
enet_selected_df <- data.frame(
  variable = rownames(enet_selected_coefs)[-1],
  coefficient = as.vector(enet_selected_coefs)[-1]
) %>%
  filter(coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

kable(enet_selected_df, digits = 2, 
      caption = "Variables Selected by Elastic Net at the Cross-Validation-Chosen Lambda") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r elastic-net-count-comparison}
cat("Number of nonzero coefficients (out of 30):\n")
cat("  Ridge:      ", sum(coef(cv_ridge, s = "lambda.min")[-1] != 0), "\n")
cat("  Lasso:      ", sum(coef(cv_lasso, s = "lambda.min")[-1] != 0), "\n")
cat("  Elastic Net:", sum(coef(cv_enet, s = "lambda.min")[-1] != 0))
```

**Comparing Variable Selection Behavior**:

- **Ridge (30/30 non-zero)**: No variable selection; retains every predictor.
- **Lasso (12/30 non-zero)**: Aggressive selection; zeroes out 18 noise variables.
- **Elastic Net (17/30 non-zero)**: Balanced selection; zeroes out 13 noise variables while offering smoother shrinkage for surviving predictors.

#### Handling Correlated Predictors: A Direct Comparison

The key advantage of Elastic Net over Lasso becomes obvious when predictors are collinear (e.g., Poverty Rate and Uninsured Rate moving together across counties).

To evaluate performance under high collinearity, we simulate two strongly correlated predictors ($r \approx 0.8$), both with an equal true effect size ($\beta = 3.0$), alongside 8 noise variables. 

To measure how each estimator behaves, we combine two complementary evaluation tools:
- **Cross-Validation (`cv.glmnet()`)**: Used *within each sample* to select the optimal tuning parameter $\lambda$ (`lambda.min`) for Ridge, Lasso, and Elastic Net.
- **Monte Carlo Simulation ($B = 100$)**: Replicates the data generation and CV tuning process across 100 independent sample realizations. This allows us to quantify the true empirical **Bias**, **Variance**, and **MSE** of each estimator across samples.

```{r elastic-net-sim, message=FALSE, warning=FALSE}
set.seed(55)

# Simulate two "true" predictors that are highly correlated with each other (r ≈ 0.8)
n_counties_sim <- 100
poverty_rate <- rnorm(n_counties_sim, 15, 5)
uninsured_rate_corr <- poverty_rate * 0.8 + rnorm(n_counties_sim, 0, 1.5)

# Both contribute equally to mortality (true effect = 3.0 each)
mortality_corr <- 500 + 3 * poverty_rate + 3 * uninsured_rate_corr + rnorm(n_counties_sim, 0, 10)

X_corr <- cbind(poverty_rate, uninsured_rate_corr,
                 matrix(rnorm(n_counties_sim * 8), nrow = n_counties_sim))
colnames(X_corr) <- c("poverty_rate", "uninsured_rate", paste0("noise", 1:8))

# Single-sample estimates for OLS, Ridge, Lasso, and Elastic Net
df_corr <- as.data.frame(X_corr)
df_corr$mortality <- mortality_corr
ols_fit <- lm(mortality ~ ., data = df_corr)
ols_result <- coef(ols_fit)[c("poverty_rate", "uninsured_rate")]

cv_ridge_corr <- cv.glmnet(X_corr, mortality_corr, alpha = 0)
cv_lasso_corr <- cv.glmnet(X_corr, mortality_corr, alpha = 1)
cv_enet_corr <- cv.glmnet(X_corr, mortality_corr, alpha = 0.5)

ridge_result <- coef(cv_ridge_corr, s = "lambda.min")[c("poverty_rate", "uninsured_rate"), ]
lasso_result <- coef(cv_lasso_corr, s = "lambda.min")[c("poverty_rate", "uninsured_rate"), ]
enet_result <- coef(cv_enet_corr, s = "lambda.min")[c("poverty_rate", "uninsured_rate"), ]

# Monte Carlo simulation (B = 100) for empirical Bias, Variance, and MSE
B <- 100
true_b <- c(3, 3)
res_ols <- matrix(0, B, 2)
res_ridge <- matrix(0, B, 2)
res_lasso <- matrix(0, B, 2)
res_enet <- matrix(0, B, 2)

for (i in 1:B) {
  p_rate <- rnorm(n_counties_sim, 15, 5)
  u_rate <- p_rate * 0.8 + rnorm(n_counties_sim, 0, 1.5)
  m_rate <- 500 + 3 * p_rate + 3 * u_rate + rnorm(n_counties_sim, 0, 10)
  Xc <- cbind(p_rate, u_rate, matrix(rnorm(n_counties_sim * 8), nrow = n_counties_sim))
  colnames(Xc) <- c("p_rate", "u_rate", paste0("noise", 1:8))
  df_i <- as.data.frame(Xc)
  df_i$m_rate <- m_rate
  
  res_ols[i, ] <- coef(lm(m_rate ~ ., data = df_i))[c("p_rate", "u_rate")]
  
  cv_r <- cv.glmnet(Xc, m_rate, alpha = 0)
  cv_l <- cv.glmnet(Xc, m_rate, alpha = 1)
  cv_e <- cv.glmnet(Xc, m_rate, alpha = 0.5)
  
  res_ridge[i, ] <- coef(cv_r, s = "lambda.min")[c("p_rate", "u_rate"), ]
  res_lasso[i, ] <- coef(cv_l, s = "lambda.min")[c("p_rate", "u_rate"), ]
  res_enet[i, ]  <- coef(cv_e, s = "lambda.min")[c("p_rate", "u_rate"), ]
}

b_ols   <- mean(abs(colMeans(res_ols) - true_b))
v_ols   <- mean(apply(res_ols, 2, var))
b_ridge <- mean(abs(colMeans(res_ridge) - true_b))
v_ridge <- mean(apply(res_ridge, 2, var))
b_lasso <- mean(abs(colMeans(res_lasso) - true_b))
v_lasso <- mean(apply(res_lasso, 2, var))
b_enet  <- mean(abs(colMeans(res_enet) - true_b))
v_enet  <- mean(apply(res_enet, 2, var))
```

```{r elastic-net-table}
corr_summary_df <- data.frame(
  Method = c("True Effect", "OLS (Unpenalized)", "Ridge (L2)", "Lasso (L1)", "Elastic Net (L1+L2)"),
  `Poverty Rate` = c(3.00, round(ols_result[1], 2), round(ridge_result[1], 2), round(lasso_result[1], 2), round(enet_result[1], 2)),
  `Uninsured Rate` = c(3.00, round(ols_result[2], 2), round(ridge_result[2], 2), round(lasso_result[2], 2), round(enet_result[2])),
  `Abs Bias` = c(0.000, round(b_ols, 3), round(b_ridge, 3), round(b_lasso, 3), round(b_enet, 3)),
  `Variance` = c(0.000, round(v_ols, 3), round(v_ridge, 3), round(v_lasso, 3), round(v_enet, 3)),
  `MSE (Bias² + Var)` = c(0.000, round(b_ols^2 + v_ols, 3), round(b_ridge^2 + v_ridge, 3), round(b_lasso^2 + v_lasso, 3), round(b_enet^2 + v_enet, 3)),
  `Behavior with Collinear Predictors` = c(
    "Ground Truth",
    "Unbiased, but high variance under collinearity",
    "Lowest variance (Grouping effect)",
    "High variance due to lopsided selection",
    "Balanced variance reduction + Selection"
  ),
  check.names = FALSE
)

kable(corr_summary_df, caption = "Empirical Comparison of Coefficients, Bias, Variance, and MSE Under High Multicollinearity (r ≈ 0.8)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r elastic-net-barchart, fig.width=11, fig.height=5}
# Left Panel: Single-Sample Estimated Coefficients
bar_df <- data.frame(
  Variable = rep(c("Poverty Rate", "Uninsured Rate"), times = 5),
  Method = rep(c("True Effect", "OLS", "Ridge", "Lasso", "Elastic Net"), each = 2),
  Coefficient = c(3.0, 3.0,
                  unname(ols_result[1]), unname(ols_result[2]),
                  unname(ridge_result[1]), unname(ridge_result[2]), 
                  unname(lasso_result[1]), unname(lasso_result[2]), 
                  unname(enet_result[1]), unname(enet_result[2]))
)
bar_df$Method <- factor(bar_df$Method, levels = c("True Effect", "OLS", "Ridge", "Lasso", "Elastic Net"))

p_coef <- ggplot(bar_df, aes(x = Variable, y = Coefficient, fill = Method)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "gray30", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "gray50") +
  scale_fill_manual(values = c("True Effect" = "#222222",
                               "OLS" = "#757575",
                               "Ridge" = "#2e7d32", 
                               "Lasso" = "#c62828", 
                               "Elastic Net" = "#1565c0")) +
  labs(title = "Single-Sample Estimates",
       subtitle = "True effect size = 3.0 (dashed line)",
       x = "", y = "Coefficient Value") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 13),
        plot.margin = ggplot2::margin(10, 10, 10, 10))

# Right Panel: Empirical Variance & Bias Trade-off Across Samples
metrics_df <- data.frame(
  Method = factor(c("OLS", "Ridge", "Lasso", "Elastic Net"), levels = c("OLS", "Ridge", "Lasso", "Elastic Net")),
  Variance = c(v_ols, v_ridge, v_lasso, v_enet),
  Bias = c(b_ols, b_ridge, b_lasso, b_enet)
)

p_metrics <- ggplot(metrics_df, aes(x = Method, y = Variance, fill = Method)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = sprintf("Var: %.3f\nBias: %.3f", Variance, Bias)), 
            vjust = -0.2, size = 3.1, fontface = "bold") +
  scale_fill_manual(values = c("OLS" = "#757575",
                               "Ridge" = "#2e7d32", 
                               "Lasso" = "#c62828", 
                               "Elastic Net" = "#1565c0")) +
  ylim(0, max(metrics_df$Variance) * 1.3) +
  labs(title = "Empirical Variance & Bias",
       subtitle = "Monte Carlo (B=100) across samples",
       x = "", y = "Variance") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 13),
        plot.margin = ggplot2::margin(10, 10, 10, 10))

p_coef + p_metrics + plot_layout(widths = c(1.3, 1))
```

**Key Empirical Findings (Bias-Variance Trade-off in Action)**:

- **True Effect (Black)**: Ground truth ($\beta = 3.0$).
- **OLS (Gray)**: Unbiased ($\text{Bias} \approx 0.091$), but suffers from high **Variance ($0.329$)** because collinear predictors cause sample-to-sample instability.
- **Ridge (Green)**: Accepts a tiny shrinkage **Bias ($0.133$)**, but slashes **Variance down to $0.055$** (an 83% reduction vs. OLS), achieving by far the lowest total MSE ($0.072$).
- **Lasso (Red)**: Has high **Variance ($0.356$)** under collinearity because it randomly flips between correlated predictors from sample to sample.
- **Elastic Net (Blue)**: Strikes a balanced compromise — cuts **Variance to $0.160$** (a 50% reduction vs. OLS/Lasso) while preserving sparse variable selection.

#### Why This Matters for Applied Research

When working with high-dimensional observational data (e.g., county health indicators), choose your regularization strategy based on your data structure:

| Situation | Reach for | Key Property |
|---|---|---|
| Many independent predictors; expect strong sparsity | **Lasso** | Aggressive variable selection |
| High multicollinearity; want to retain all controls | **Ridge** | Stable $L_2$ shrinkage across all features |
| Many correlated predictors; want sparse yet stable selection | **Elastic Net** | Grouping effect + variable selection |

> **Important Caution**: Regularized coefficients are intentionally biased ($\hat{\beta}_{\text{reg}} \to 0$) to reduce variance. They excel at prediction and feature selection, but should **not** be interpreted directly as unbiased causal effect sizes. For causal inference, use Lasso or Elastic Net to select control variables, then estimate an unpenalized model (e.g., Fixed Effects).

### Conclusion

Supervised regularization links four core concepts into a cohesive workflow:

1. **Supervised Framework**: Learning from labeled outcomes ($Y$) to predict new data.
2. **Two Cultures**: Prioritizing out-of-sample prediction accuracy ($\hat{Y}$) over in-sample coefficient interpretation ($\hat{\beta}$).
3. **Bias-Variance Tradeoff**: Balancing model capacity to avoid overfitting training noise.
4. **Regularization**: Using $\lambda$ penalties (Ridge, Lasso, Elastic Net) to control complexity systematically.

In the next post, we will explore **tree-based methods** (Decision Trees, Random Forests, Gradient Boosting) to see how non-linear complexity and variance reduction operate in non-parametric models.
